Online {\IterComp}

### 2017-09-12 cgo18mscr.tex
minor changes in results
updates motivation and other minor changes

\maketitle

@@

In contrast to the naive instrumentation where each basic block records only its own amount of work, with the optimal profiling, the instrumented basic blocks need record an aggregated value of work that represents a path in the CFG.
These values are constructed with some instrumented basic blocks speculatively assuming some paths while other probes correct when these assumptions are wrong (see for example $w(P_0)$ and $w(P_1)$ in Figure~\ref{fig:cfg-example}).

\begin{lstlisting}[caption={Post-processing of the CFG for populating all edge flows based on the collected probes.
\textbf{ToDo: Replace the algorithm with a formal data-flow equation.}}, label={lst:populateEdgeFlows}, float]
// Inputs: CFG with the known edges flows from
//         the cotree (collected probes).
// Output: Updated CFG with all edge flows.
populateEdgeFlows(G) {
  changed = true
  while changed:
    changed = false
    for B in G.vertices():
      unIN = count( G.unknownIncomingEdges(B) )
      unOUT = count( G.unknownOutgoingEdges(B) )
      if unIN==0 and unOUT==1:
        //sum known incoming and outgoing edges
        sIN = sum( G.incomingEdges(B) )
        sOUT = sum( G.outgoingEdges(B) )
        //update unknown outgoing edge in B
        //with (sIN-sOUT)
        G.setUnknownOutgoingEdge(B, (sIN-sOUT))
        changed = true
      if unIN==1 and unOUT==0:
        //sum known incoming and outgoing edges
        sIN = sum( G.incomingEdges(B) )
        sOUT = sum( G.outgoingEdges(B) )
        //update unknown incoming edge in B
        //with (sOUT-sIN)
        G.setUnknownIncomingEdge(B, (sOUT-sIN))
        changed = true
}
\end{lstlisting}

Because the algorithm for the optimal placement of the probes is proved to uniquely compute the block frequencies by propagating the probe counts, we adapt this algorithm in order to compose the aggregated values that will be instrumented in each probe, based on our model of computational work, $\Delta W$, derived from the basic block frequencies (see Section~\ref{sec:metric}).
We perform a similar propagation of the probes in a symbolic fashion, as illustrated in Figure~\ref{fig:cfg-example}.
This symbolic propagation of the probes is implemented by the data-flow analysis described in Algorithm~\ref{alg:populateEdgeInfo}, and the final aggregated values are extracted from the edge information as described by Algorithm~\ref{alg:instrValue}.

The data-flow analysis in Algorithm~\ref{alg:populateEdgeInfo} keeps two sets for each edge, namely, the increment and the decrement sets.
We consider that both sets represent the \textit{edge expressions} shown in Figure~\ref{fig:cfg-example}, for which we define a \textit{symbolic sum} by computing the union of the increment and decrement sets, respectively, with the appropriate cancellation of common elements.
This data-flow analysis is based on the invariant that the symbolic sum of all the incoming edges must equals the symbolic sum of the outgoing edges.
For example, the symbolic sum of the incoming edges of the basic block $B_8$ is $P_0 - P_1$, where $P_3$ is cancelled out.

@@

Algorithm~\ref{alg:instrValue} reads the edge information for each basic block by computing the symbolic sum of their respective incoming edges (or outgoing edges).
From these \textit{edge expressions}, we are able to compose the aggregated value of the probes.
The positive terms in the edge expression of a basic block indicate that the amount of work of this basic block will be incremented in the probes represented by these positive terms, similarly, the negative terms indicate that the amount of work of this basic block will be decremented in the probes represented by these negative terms.
For example, because the edge expression for the basic block $B_8$ is $P_0 - P_1$, the amount of work of $B_8$, denoted by $w(B_8)$, is incremented in probe $P_0$ and decremented in $P_1$.

\begin{lstlisting}[caption={Pseudocode that describes how the edge information is used in order to extract the value that will be computed in a given instrumented basic block $B_I$. This algorithm could equally be implemented based on the predecessors.
\textbf{ToDo: Using the vector format, we could replace this algorithm with a single equation. Let $V$ be the vector of $\{-1,0,1\}$ and $W$ the weight for each basic block.
This algorithm could be re-written as $reduce (+) (map (*) (zip V W))$.}
}, label={lst:instrValue}, float]
// Inputs: 1) CFG with the known edges flows
//            of the chords.
//         2) Basic block targeted for probing.
// Output: Work value to be incremented by the
//         given probe B.
instrValue(G, B, inc, dec) {
  value = 0
  for B in G.vertices():
    incB = set()
    decB = set()
    for succB in G.successor(B):
      incB = incB union inc[G.getEdge(B, succB)]
      decB = decB union dec[G.getEdge(B, succB)]
    if B in (incB - decB):
      value = value + w(B)
    if B in (decB - incB):
      value = value - w(B)
  return value
}
\end{lstlisting}

@@

%Figure~\ref{fig:relax-instr-diagram} shows an overview of the relaxed instrumentation algorithm.
%The highlighted step is introduced by the relaxed instrumentation on top of the previously defined optimal profiling.
The relaxed instrumentation performs a post-processing on the resulting instrumentation of the optimal algorithm.
This post-processing identifies probes that add little to the work metric, and removes their instrumentation.
In order to guarantee an upper bound for the dynamic error in the profiling measurement, the relaxation algorithm applies a constrained post-processing on a per DAG (directed acyclic graph) basis.
By constraining the relaxation within each DAG by a maximum allowed static error, it guarantees that the overall relaxation will also be constrained by the same bound.

>>>>

\maketitle
\vspace{-1em}

@@

In contrast to the naive instrumentation where each basic block records only its own amount of work, with the optimal profiling, the instrumented basic blocks need to record an aggregated value of work that represents a path in the CFG.
These values are constructed with some instrumented basic blocks speculatively assuming some paths while other probes correct when these assumptions are wrong (see for example $w(P_0)$ and $w(P_1)$ in Figure~\ref{fig:cfg-example}).

\begin{lstlisting}[caption={Pseudocode of the data-flow analysis necessary for assigning the values computed in each probe.},
label={lst:populateEdgeInfo}, float]
// Inputs: CFG with the known edges flows from
//         the cotree (collected probes).
// Output: Updated CFG with all edge flows.
populateEdgeFlows(G) {
  changed = true
  while changed:
    changed = false
    for B in G.vertices():
      unIN = count( G.unknownIncomingEdges(B) )
      unOUT = count( G.unknownOutgoingEdges(B) )
      if unIN==0 and unOUT==1:
        //sum known incoming and outgoing edges
        sIN = sum( G.incomingEdges(B) )
        sOUT = sum( G.outgoingEdges(B) )
        //update unknown outgoing edge in B
        //with (sIN-sOUT)
        G.setUnknownOutgoingEdge(B, (sIN-sOUT))
        changed = true
      if unIN==1 and unOUT==0:
        //sum known incoming and outgoing edges
        sIN = sum( G.incomingEdges(B) )
        sOUT = sum( G.outgoingEdges(B) )
        //update unknown incoming edge in B
        //with (sOUT-sIN)
        G.setUnknownIncomingEdge(B, (sOUT-sIN))
        changed = true
}
\end{lstlisting}

Because the algorithm for the optimal placement of the probes is proved to uniquely compute the block frequencies by propagating the probe counts, we adapt this algorithm (see \lstlistingname~\ref{lst:populateEdgeFlows}) in order to compose the aggregated values that will be instrumented in each probe, based on our model of computational work, $\Delta W$, derived from the basic block frequencies (see Section~\ref{sec:metric}).
We perform a similar propagation of the probes in a symbolic fashion, as illustrated in Figure~\ref{fig:cfg-example}.
This symbolic propagation of the probes is implemented by the data-flow analysis described in \lstlistingname~\ref{lst:populateEdgeInfo}, and the final aggregated values are extracted from the edge information as described by \lstlistingname~\ref{lst:instrValue}.

The data-flow analysis in \lstlistingname~\ref{lst:populateEdgeInfo} keeps two sets for each edge, namely, the increment and the decrement sets.
We consider that both sets represent the \textit{edge expressions} shown in Figure~\ref{fig:cfg-example}, for which we define a \textit{symbolic sum} by computing the union of the increment and decrement sets, respectively, with the appropriate cancellation of common elements.
This data-flow analysis is based on the invariant that the symbolic sum of all the incoming edges must equals the symbolic sum of the outgoing edges.
For example, the symbolic sum of the incoming edges of the basic block $B_8$ is $P_0 - P_1$, where $P_3$ is cancelled out.

@@

\lstlistingname~\ref{lst:instrValue} reads the edge information for each basic block by computing the symbolic sum of their respective incoming edges (or outgoing edges).
From these \textit{edge expressions}, we are able to compose the aggregated value of the probes.
The positive terms in the edge expression of a basic block indicate that the amount of work of this basic block will be incremented in the probes represented by these positive terms. Similarly, the negative terms indicate that the amount of work of this basic block will be decremented in the probes represented by these negative terms.
For example, because the edge expression for the basic block $B_8$ is $P_0 - P_1$, the amount of work of $B_8$, denoted by $w(B_8)$, is incremented in probe $P_0$ and decremented in $P_1$.

%Let $V$ be a vector of $\{-1,0,1\}$ with the sum of the edge information from the incoming edges of a given basic block.
%Let $W$ be the 
%\[
%\mathrm{reduce}\,\,(+)\,\,(\mathrm{map}\,\,(\times W)\,\,E^+(B))
%\]

\begin{lstlisting}[caption={Pseudocode that describes how the edge information is used in order to extract the value that will be computed in a given instrumented basic block $B_I$. This algorithm could equally be implemented based on the outgoing edges.}, label={lst:instrValue}, float]
// Inputs: 1) CFG with the known edges flows
//            of the chords.
//         2) Probe to be instrumented.
// Output: Work value to be incremented by the
//         given probe P.
instrValue(G, P) {
  value = 0
  for B in G.vertices():
    expr = sum( G.incomingEdges(B) ) 
    //coef(P,expr) is the coefficient
    //-1,0, or 1 of P in expr
    value = value + coef(P, expr)*w(B)
  return value
}
\end{lstlisting}

@@

%Figure~\ref{fig:relax-instr-diagram} shows an overview of the relaxed instrumentation algorithm.
%The highlighted step is introduced by the relaxed instrumentation on top of the previously defined optimal profiling.
The relaxed instrumentation performs a post processing on the resulting instrumentation of the optimal algorithm.
This post-processing identifies probes that add little to the work metric, and removes their instrumentation.
In order to guarantee an upper bound for the dynamic error in the profiling measurement, the relaxation algorithm applies a constrained post-processing on a per DAG (directed acyclic graph) basis.
By constraining the relaxation within each DAG by a maximum allowed static error, it guarantees that the overall relaxation will also be constrained by the same bound.

### 2017-09-12 motivation.tex
updates motivation and other minor changes

However, because of the restriction of having a single execution per input,
it is not possible to measure speedup in a real-world online scenario.
Moreover, measuring just execution time, for example, is also not viable since
different inputs often mean different amounts of work,
rendering it meaningless to compare optimizations only by execution time.

@@

In the remaining of this paper we develop a work efficiency metric, suitable
for guiding online {\itercomp}, which can be profiled with a single execution
of the program with the optimization sequence being evaluated.

>>>>

However, because of the restriction of having a single execution per input,
it is not possible to measure speedup in a real-world online scenario.
Moreover, measuring just execution time, for example, is also not viable since
different inputs often mean different amounts of work,
rendering it meaningless to compare optimizations only by execution time.
Therefore, we need a performance metric which can be profiled from a single
execution of the program.
Although a previous work has suggested that \textit{instructions per cycle}~(IPC)
could be used with that goal, IPC have no correlation with speedup~\citep{fursin07}.
In a case where we have two distinct versions of the program, due to different
optimizations, which is also executed with distinct inputs, higher IPC does not
necessarily mean lower execution time (higher performance).

@@

Figure~\ref{fig:motivation-metric} shows a comparison among these three metrics,
namely speedup, execution time, and IPC, and their correlation with the actual
speedup.
Each point in the figure corresponds to each metric averaged over 
a set of inputs, collected during the {\itercomp}.
The vertical lines correspond to the optimization sequence 
with the best score for each one of the metrics.
This figure shows that both execution time and IPC are unsuitable for
online {\itercomp}, as they present no direct correlation with the actual
speedup obtained by the optimization sequence being evaluated.

In the remaining of this paper, we develop a work efficiency metric, suitable
for guiding online {\itercomp}, which can be profiled with a single execution
of the program with the optimization sequence being evaluated.

### 2017-09-12 results.tex
updates to motivation
updates motivation and other minor changes

@@

\begin{figure*}[ht]
    \centering
    \includegraphics[width=\textwidth]{figs/overhead-O3.pdf}
    \caption{Overhead of the instrumentations compiled with {\flagstype -O3}.}
    \label{fig:overhead-O3}
\end{figure*}

@@

\begin{figure*}[h!]
    \centering
    \includegraphics[width=\textwidth]{figs/error-O3.pdf}
    \caption{Dynamic error of the work profiling averaged over the 1000 inputs, after relaxing the number of probes.}
    \label{fig:error-O3}
\end{figure*}

@@

For comparison, we use four configurations\footnote{A configuration using the WPO relaxation was not used due to the time necessary for executing the experiments.}.
In all configurations, the same optimization sequence is used for multiple inputs, using a dynamic input-window size, as explained in Section~\ref{sec:oic-infra}.
The average performance over the input window provides an estimate for the overall performance of the optimization sequence across distinct inputs.
The optimization sequences are ranked based on their average performance, and the best optimization sequence is selected.
%When selecting the best optimization sequence, they are ranked by the average of their performance measurement, using the WP metric, except for the Oracle-RM which uses the real speedup over \texttt{-O0}.
\begin{itemize}
\item \textbf{Oracle-RM} executes the program twice for each input, measuring the real speedup for each optimization sequence, and then uses the real speedup over {\flagstype -O0} for guiding online {\itercomp}. This oracle provides a reference results for assessing the effectiveness work-efficiency metric.
\item \textbf{Oracle-PP} represents an oracle with a \textit{perfect} non-intrusive profiling.
  It uses the work-efficiency metric for guiding online {\itercomp}.
  Because this version simulates a \textit{perfect} profiling, with zero overhead,
  it allows us isolate and evaluate the effectiveness work-efficiency metric.
\item \textbf{Real-OP} corresponds to the online {\itercomp} as it would be applied in a real-world online scenario.
  The program is executed only once with each input.
  It uses the optimal work profiling for computing the work efficiency.
\item \textbf{Real-R5} is similar to the {Real-OP}, but with a relaxation of 5\% in the work profiling.
\item \textbf{Real-IPC} uses IPC for guiding the online {\itercomp}.
\end{itemize}
The comparison between Oracle-RM and Oracle-PP is useful for validating the use of the WP metric for guiding {\itercomp}, while the other configurations demonstrate the viability of applying the online {\itercomp} in a real scenario.

@@

\begin{figure*}[htb]
    \centering
    \includegraphics[width=\textwidth]{figs/window-size.pdf}
    \caption{Average input-window sizes observed during the online {\itercomp}.}
    \label{fig:window-size}
\end{figure*}

@@

Repeating the same optimization pass can be beneficial and usually expected by other passes.
For example, the {\flagstype -loop-simplify} pass is used for transforming loops into a canonical form by inserting pre-header and exit basic blocks.
Although this pass inserts jumps due to redundant basic blocks, this canonical form can be favourable to other loop optimizations.
Because of the redundant basic blocks, this optimization pass expects that the {\flagstype -simplifycfg} will eventually be executor later on the optimization pipeline.
Another example of such inter-relation between transformations concerns the {\flagstype -licm} and {\flagstype -mem2reg} passes.
The {\flagstype -licm} pass is responsible for moving invariant code out from the loop body.
It usually creates new local variables, using memory access operations, for assisting with the code manipulation, which means that the executing the {\flagstype -mem2reg} pass afterwards would be useful as a cleanup pass for removing the extra memory accesses generated.
However, many of the analysis required for identifying loop invariant also benefit from the transformations performed by the {\flagstype -mem2reg} pass.
These examples illustrate the importance of repeating optimization passes.
Moreover, they illustrate the intricate relation amongst several transformations.

However, all optimization sequences in the set of optimizations were generated completely at random, without using any knowledge of individual transformations.
Each optimization sequence was generated in two steps: \textit{(1)} randomly selects the number of flags; \textit{(2)} randomly selects the flags, allowing repetitions.
Afterwards, this randomly generated optimization sequence would be included in the set of optimization sequences only if it was able to improve the performance of a training benchmark, also selected at random, in respect of the {\flagstype -O3} default optimization.
This process was repeated until we obtained all the 500 distinct optimization sequences.

@@

In order to evaluate the quality of the final optimization sequences selected by the {\itercomp} search, we compare their speedup by measuring wall-clock time of the benchmarks when compiled with the standard {\flagstype -O3} optimization.
For each benchmark, after selecting the final optimization sequence, we compute the average speedup over all the 1000 input dataset.
When measuring the wall-clock time for each input, to reduce noise, we execute the same input until we have a statistically sound measurement, i.e. we execute until we have an interval no larger than 1\% with 99\% confidence.
Figure~\ref{fig:speedups} shows these average speedups over all the 1000 input dataset.
This figure shows that the best optimization sequence selected with the Oracle-PP is very close to the performance of the best optimization sequence selected with the Oracle-RM.
This result is important for demonstrating that a work-based metric has the potential to produce good results in a real online scenario, where there is the restriction that programs execute distinct inputs only once.
Moreover, the evaluation also indicates that the use of relaxation algorithms does not degrade the optimization search,
it might even benefit the {\itercomp} as it tends to insert smaller interferences in the performance measurement.
Finally, Figure~\ref{fig:speedups} also presents the results of {\itercomp} guided by IPC, which is futher discussed in Figure~\ref{fig:ipc-vs-work}.

\begin{figure*}[htb]
    \centering
    \includegraphics[width=\textwidth]{figs/speedups.pdf}
    \caption{Speedups obtained from the final optimization sequence selected by the online {\itercomp}.
	         The speedups reported for each benchmark represents the average speedup across their complete 1000 input datasets.}
    \label{fig:speedups}
\end{figure*}

@@

\textbf{ToDo: Include results using Genetic Algorithm search.}

>>>>

\begin{figure*}[t]
    \centering
    \includegraphics[width=\textwidth]{figs/overhead-O3.pdf}
    \caption{Overhead of the instrumentations compiled with {\flagstype -O3}.}
    \label{fig:overhead-O3}
\end{figure*}

\begin{figure*}[t]
    \centering
    \includegraphics[width=\textwidth]{figs/error-O3.pdf}
    \caption{Dynamic error of the work profiling averaged over the 1000 inputs, after relaxing the number of probes.}
    \label{fig:error-O3}
\end{figure*}

@@

@@

@@

For comparison, we use four configurations\footnote{A configuration using the WPO relaxation was not used due to the time necessary for executing the experiments.}.
In all configurations, the same optimization sequence is used for multiple inputs, using a dynamic input-window size, as explained in Section~\ref{sec:oic-infra}.
The average performance over the input window provides an estimate for the overall performance of the optimization sequence across distinct inputs.
The optimization sequences are ranked based on their average performance, and the best optimization sequence is selected.
%When selecting the best optimization sequence, they are ranked by the average of their performance measurement, using the WP metric, except for the Oracle-RM which uses the real speedup over \texttt{-O0}.
\begin{itemize}
\item \textbf{Oracle-RM} executes the program twice for each input, measuring the real speedup for each optimization sequence, and then uses the real speedup over {\flagstype -O0} for guiding online {\itercomp}. This oracle provides a reference results for assessing the effectiveness work-efficiency metric.
\item \textbf{Oracle-PP} represents an oracle with a \textit{perfect} non-intrusive profiling.
  It uses the work-efficiency metric for guiding online {\itercomp}.
  Because this version simulates a \textit{perfect} profiling, with zero overhead,
  it allows us isolate and evaluate the effectiveness work-efficiency metric.
\item \textbf{Real-OP} corresponds to the online {\itercomp} as it would be applied in a real-world online scenario.
  The program is executed only once with each input.
  It uses the optimal work profiling for computing the work efficiency.
\item \textbf{Real-R5} is similar to the {Real-OP}, but with a relaxation of 5\% in the work profiling.
\item \textbf{Real-WPO-R5} is similar to the {Real-OP}, but with a whole-program relaxation of 5\% in the work profiling.
%\item \textbf{Real-IPC} uses IPC for guiding the online {\itercomp}.
\end{itemize}
The comparison between Oracle-RM and Oracle-PP is useful for validating the use of the WP metric for guiding {\itercomp}, while the other configurations demonstrate the viability of applying the online {\itercomp} in a real scenario.

@@

\begin{figure*}[t]
    \centering
    \includegraphics[width=\textwidth]{figs/window-size.pdf}
    \caption{Average input-window sizes observed during the online {\itercomp}.}
    \label{fig:window-size}
\end{figure*}

@@

%Repeating the same optimization pass can be beneficial and usually expected by other passes.
%For example, the {\flagstype -loop-simplify} pass is used for transforming loops into a canonical form by inserting pre-header and exit basic blocks.
%Although this pass inserts jumps due to redundant basic blocks, this canonical form can be favourable to other loop optimizations.
%Because of the redundant basic blocks, this optimization pass expects that the {\flagstype -simplifycfg} will eventually be executor later on the optimization pipeline.
%Another example of such inter-relation between transformations concerns the {\flagstype -licm} and {\flagstype -mem2reg} passes.
%The {\flagstype -licm} pass is responsible for moving invariant code out from the loop body.
%It usually creates new local variables, using memory access operations, for assisting with the code manipulation, which means that the executing the {\flagstype -mem2reg} pass afterwards would be useful as a cleanup pass for removing the extra memory accesses generated.
%However, many of the analysis required for identifying loop invariant also benefit from the transformations performed by the {\flagstype -mem2reg} pass.
%These examples illustrate the importance of repeating optimization passes.
%Moreover, they illustrate the intricate relation amongst several transformations.

All optimization sequences in the set of optimizations were generated completely at random, without using any knowledge of individual transformations.
Each optimization sequence was generated in two steps: \textit{(1)} randomly selects the number of flags; \textit{(2)} randomly selects the flags, allowing repetitions.
Afterwards, this randomly generated optimization sequence would be included in the set of optimization sequences only if it was able to improve the performance of a training benchmark, also selected at random, in respect of the {\flagstype -O3} default optimization.
This process was repeated until we obtained all the 500 distinct optimization sequences.

@@

In order to evaluate the quality of the final optimization sequences selected by the {\itercomp} search, we compare their speedup by measuring wall-clock time of the benchmarks when compiled with the standard {\flagstype -O3} optimization.
For each benchmark, after selecting the final optimization sequence, we compute the average speedup over all the 1000 input dataset.
When measuring the wall-clock time for each input, to reduce noise, we execute the same input until we have a statistically sound measurement, i.e. we execute until we have an interval no larger than 1\% with 99\% confidence.
Figure~\ref{fig:speedups} shows these average speedups over all the 1000 input dataset.
This figure shows that the best optimization sequence selected with the Oracle-PP is very close to the performance of the best optimization sequence selected with the Oracle-RM.
This result is important for demonstrating that a work-based metric has the potential to produce good results in a real online scenario, where there is the restriction that programs execute distinct inputs only once.
Moreover, the evaluation also indicates that the use of relaxation algorithms does not degrade the optimization search,
it might even benefit the {\itercomp} as it tends to insert smaller interferences in the performance measurement.
%Finally, Figure~\ref{fig:speedups} also presents the results of {\itercomp} guided by IPC, which is futher discussed in Figure~\ref{fig:ipc-vs-work}.

\begin{figure*}[t]
    \centering
    \includegraphics[width=\textwidth]{figs/speedups.pdf}
    \caption{Speedups obtained from the final optimization sequence selected by the online {\itercomp}.
	         The speedups reported for each benchmark represents the average speedup across their complete 1000 input datasets.}
    \label{fig:speedups}
\end{figure*}

@@

%\textbf{ToDo: Include results using Genetic Algorithm search.}
